%==========================================================
%  Section 3: Implementation
%==========================================================
\section{Implementation}

%----------------------------------------------------------
\subsection{Feature Engineering via Grouped PCA}
%----------------------------------------------------------

\subsubsection{Motivation}

The raw training dataset contains several hundred meteorological predictors,
many of which share strong inter-group collinearity.
For instance, the NMME ensemble model outputs for temperature forecasts
(\texttt{nmme-tmp2m-34w}, \texttt{nmme0-tmp2m-34w}) and geopotential height fields
at multiple pressure levels (\texttt{hgt10}, \texttt{hgt100}, \texttt{hgt500},
\texttt{hgt850}) each form a block of near-redundant variables.
Feeding all raw features directly into a regression model risks overfitting,
inflates training cost, and may obscure the dominant climate signal
within each physical group.
We therefore apply \textbf{Grouped PCA} to compress each meteorological block
independently while preserving its principal modes of variability.

\subsubsection{Principal Component Analysis}

Let $\mathbf{X} \in \mathbb{R}^{n \times p}$ denote a mean-centred, unit-variance
scaled feature matrix with $n$ samples and $p$ variables.
PCA seeks an orthogonal transformation
$\mathbf{W} = [\mathbf{w}_1, \ldots, \mathbf{w}_k] \in \mathbb{R}^{p \times k}$
that maximises the variance of the projected data:

\begin{equation}
  \mathbf{w}_j = \underset{\|\mathbf{w}\|=1}{\arg\max}\;
  \mathrm{Var}(\mathbf{X}\mathbf{w})
  \quad \text{s.t.}\quad \mathbf{w}_j \perp \mathbf{w}_i,\; \forall\, i < j.
  \label{eq:pca}
\end{equation}

The solution is given by the $k$ leading eigenvectors of the sample covariance
matrix $\mathbf{\Sigma} = \frac{1}{n-1}\mathbf{X}^\top\mathbf{X}$,
equivalently the top-$k$ right singular vectors from the SVD decomposition
$\mathbf{X} = \mathbf{U}\mathbf{\Sigma}\mathbf{V}^\top$.
The resulting score matrix
$\mathbf{Z} = \mathbf{X}\mathbf{W} \in \mathbb{R}^{n \times k}$
contains $k$ uncorrelated principal components (PCs) that capture the maximum
variance in the original data.

\subsubsection{Grouped PCA Design}

Rather than applying a single global PCA, we partition the feature space into
11 meteorologically meaningful groups and fit an independent
PCA within each group on the \emph{raw} (unscaled) features
(Table~\ref{tab:pca_groups}).
For each group, the number of principal components is chosen dynamically
as the smallest $k$ such that the cumulative explained variance ratio
$\sum_{j=1}^{k}\lambda_j / \sum_{j=1}^{p}\lambda_j \geq 0.95$,
ensuring that at least 95\% of the within-group variance is preserved.
This design (i) respects the physical boundaries between variable types,
(ii) prevents dominant groups from absorbing the variance budget of smaller groups,
and (iii) adaptively reduces the total feature dimension from $\sim$500+ to 92
(61 PCA components $+$ 31 pass-through features)
whilst retaining interpretable PC names.
Features not belonging to any PCA group (e.g.\ latitude, longitude,
temporal encodings, \texttt{contest-*} benchmark predictors) are passed
through unchanged.

\begin{table}[htbp]
  \centering
  \caption{Grouped PCA configuration (95\% variance threshold).}
  \label{tab:pca_groups}
  \begin{tabular}{llcc}
    \toprule
    Group name & Source variables & \#\ Input cols & \#\ PCs \\
    \midrule
    \texttt{nmme\_tmp34w} & NMME 3--4 week temperature ensemble & varies & 2 \\
    \texttt{nmme\_tmp56w} & NMME 5--6 week temperature ensemble & varies & 1 \\
    \texttt{nmme\_prate}  & NMME precipitation ensemble (3--6 wk) & varies & 12 \\
    \texttt{hgt10}        & Geopotential height at 10 hPa  & varies & 2 \\
    \texttt{hgt100}       & Geopotential height at 100 hPa & varies & 2 \\
    \texttt{hgt500}       & Geopotential height at 500 hPa & varies & 6 \\
    \texttt{hgt850}       & Geopotential height at 850 hPa & varies & 8 \\
    \texttt{sst}          & Sea surface temperature (2010--) & varies & 2 \\
    \texttt{icec}         & Sea ice concentration (2010--)   & varies & 3 \\
    \texttt{wind\_uv}     & U/V wind at 250/925 hPa          & varies & 21 \\
    \texttt{tele}         & Teleconnection indices (MEI, MJO) & varies & 2 \\
    \midrule
    \multicolumn{3}{l}{\textbf{Total PCA components}} & \textbf{61} \\
    \bottomrule
  \end{tabular}
\end{table}

The PCA is fitted \emph{exclusively on the training split} and
applied without re-fitting to the validation and test sets, ensuring no
look-ahead leakage.
No feature-wise standardisation is applied prior to PCA;
because features within each meteorological group share similar physical
scales, raw-scale PCA naturally preserves the dominant variance directions
without artificially equalising low-variance noise channels.

%----------------------------------------------------------
\subsection{Predictive Models}
%----------------------------------------------------------

All models share the same input feature set produced by Grouped PCA
(92-dimensional output) and are wrapped in a
\texttt{sklearn.pipeline.Pipeline} that prepends a \texttt{StandardScaler}
to normalise inputs before tree-based methods (for numerical stability)
and to satisfy the assumptions of linear methods.
Hyperparameters are selected via \textbf{GridSearchCV} with
\textbf{TimeSeriesSplit} ($n\_splits=5$) to respect temporal ordering.

\subsubsection{Lasso Regression}

Lasso augments the ordinary least-squares objective with an $\ell_1$ penalty:

\begin{equation}
  \hat{\boldsymbol{\beta}}_{\text{Lasso}} =
  \underset{\boldsymbol{\beta}}{\arg\min}\;
  \frac{1}{n}\|\mathbf{y} - \mathbf{X}\boldsymbol{\beta}\|_2^2
  + \alpha \|\boldsymbol{\beta}\|_1.
  \label{eq:lasso}
\end{equation}

The $\ell_1$ norm induces sparsity by driving irrelevant coefficients exactly
to zero, effectively performing automatic feature selection.
We set the regularisation strength to $\alpha = \text{\texttt{LASSO\_ALPHA}}$
and allow up to 5000 coordinate-descent iterations.

\subsubsection{Elastic Net}

Elastic Net combines the $\ell_1$ and $\ell_2$ penalties,
balancing sparsity and grouping effects:

\begin{equation}
  \hat{\boldsymbol{\beta}}_{\text{EN}} =
  \underset{\boldsymbol{\beta}}{\arg\min}\;
  \frac{1}{n}\|\mathbf{y} - \mathbf{X}\boldsymbol{\beta}\|_2^2
  + \alpha\!\left[\rho\|\boldsymbol{\beta}\|_1
    + \frac{1-\rho}{2}\|\boldsymbol{\beta}\|_2^2\right],
  \label{eq:elasticnet}
\end{equation}

where $\rho \in [0,1]$ is the $\ell_1$ mixing ratio (\texttt{l1\_ratio}).
We use $\alpha = \text{\texttt{ENET\_ALPHA}}$ and $\rho = \text{\texttt{ENET\_L1\_RATIO}}$.

\subsubsection{Random Forest}

Random Forest constructs an ensemble of $B$ decision trees
$\{T_b\}_{b=1}^B$, each grown on a bootstrap replicate of the training data
and using a random subset of $\lfloor\sqrt{p}\rfloor$ features at every split
(CART criterion for regression):

\begin{equation}
  \hat{y} = \frac{1}{B}\sum_{b=1}^{B} T_b(\mathbf{x}).
  \label{eq:rf}
\end{equation}

The randomisation over both samples and features decorrelates the trees,
reducing variance without substantially increasing bias.
We use $B=50$ trees with \texttt{max\_depth}$=10$;
\texttt{max\_depth} is further tuned over $\{6,8,10,12,14\}$ and
\texttt{min\_samples\_leaf} over $\{5,10\}$ via grid search.

\subsubsection{XGBoost}

XGBoost (eXtreme Gradient Boosting) builds an additive ensemble
of regression trees in a stage-wise fashion by minimising a regularised
second-order Taylor expansion of the loss:

\begin{equation}
  \mathcal{L}^{(t)} = \sum_{i=1}^{n}
  \left[g_i f_t(\mathbf{x}_i) + \tfrac{1}{2}h_i f_t^2(\mathbf{x}_i)\right]
  + \Omega(f_t),
  \label{eq:xgboost}
\end{equation}

where $g_i = \partial_{\hat{y}^{(t-1)}}\ell(y_i, \hat{y}^{(t-1)})$ and
$h_i = \partial^2_{\hat{y}^{(t-1)}}\ell(y_i, \hat{y}^{(t-1)})$ are the first
and second gradients of the MSE loss, and
$\Omega(f_t) = \gamma T + \tfrac{1}{2}\lambda\|\mathbf{w}\|^2$ penalises tree
complexity ($T$: leaf count, $\mathbf{w}$: leaf weights) \cite{Chen2016}.
Configuration: \texttt{n\_estimators}$=1000$, \texttt{max\_depth}$=6$,
\texttt{learning\_rate}$=0.1$; grid search tunes
\texttt{max\_depth} $\in\{4,6,8\}$ and \texttt{learning\_rate} $\in\{0.03,0.05,0.1,0.2\}$.

\subsubsection{LightGBM}

LightGBM shares the GBDT objective of XGBoost but uses two key
algorithmic innovations for scalability:
\textit{Gradient-based One-Side Sampling} (GOSS), which retains all instances
with large gradients while randomly sub-sampling those with small gradients,
and \textit{Exclusive Feature Bundling} (EFB), which merges mutually exclusive
sparse features to reduce the effective feature dimension.
Furthermore, LightGBM grows trees \emph{leaf-wise} rather than level-wise,
enabling deeper effective trees with fewer splits.
The leaf count \texttt{num\_leaves}$=31$ (tuned over $\{15,23,31\}$) controls
model complexity more directly than depth alone;
\texttt{learning\_rate} is tuned over $\{0.03,0.05,0.1,0.2\}$.

\subsubsection{CatBoost}

CatBoost introduces \textit{Ordered Boosting} to
eliminate the target leakage inherent in standard GBDT: for each training
example, the model that generated its gradient is trained on a permutation
of the dataset that does not include that example.
It also implements an efficient ordered target statistics encoding for
categorical variables.
Configuration: \texttt{iterations}$=1000$, \texttt{depth}$=6$,
\texttt{learning\_rate}$=0.1$, \texttt{early\_stopping\_rounds}$=50$.
Grid search tunes \texttt{depth} $\in\{4,6,8\}$,
\texttt{learning\_rate} $\in\{0.01,0.03,0.05,0.1\}$,
\texttt{l2\_leaf\_reg} $\in\{1,3,5,7,9\}$, and
\texttt{bagging\_temperature} $\in\{0,1\}$.

%----------------------------------------------------------
\subsection{Stacking Ensemble}
%----------------------------------------------------------

To combine the strengths of all base learners we employ
\textbf{Stacked Generalisation} (Stacking).
The procedure is:

\begin{enumerate}
  \item Select the top-$K=3$ base models ranked by their reconstructed test RMSE.
  \item Generate \emph{out-of-fold} (OOF) predictions for each base model using
        \textbf{TimeSeriesSplit} ($n\_splits=5$) on the training set,
        producing a meta-feature matrix
        $\tilde{\mathbf{X}} \in \mathbb{R}^{n_\text{train} \times K}$.
  \item Fit five candidate meta-fusion strategies on the OOF matrix:
        (a) non-negative convex combination via SLSQP ($\sum w_i=1$),
        (b) inverse-OOF-RMSE$^2$ weighting,
        (c) RidgeCV with inner time-series CV,
        (d) non-negative Lasso, and
        (e) OLS baseline.
        The strategy with the lowest OOF RMSE is selected automatically.
  \item At inference, obtain base-model predictions on the test set and
        feed them to the chosen meta-learner to produce the final forecast.
\end{enumerate}

Using OOF predictions prevents the meta-learner from seeing test information
during training, and respecting temporal ordering in the CV splits avoids
look-ahead bias.

%==========================================================
%  Section 4: Evaluation
%==========================================================
\section{Evaluation}

%----------------------------------------------------------
\subsection{Evaluation Metrics}
%----------------------------------------------------------

Model performance is quantified with two complementary metrics.

\paragraph{Root Mean Squared Error (RMSE).}
\begin{equation}
  \text{RMSE} = \sqrt{\frac{1}{n}\sum_{i=1}^{n}(y_i - \hat{y}_i)^2}.
  \label{eq:rmse}
\end{equation}
RMSE is expressed in the same unit as the target variable (°F for temperature)
and penalises large errors more heavily than small ones.

\paragraph{Coefficient of Determination ($R^2$).}
\begin{equation}
  R^2 = 1 - \frac{\sum_{i=1}^{n}(y_i - \hat{y}_i)^2}
                  {\sum_{i=1}^{n}(y_i - \bar{y})^2},
  \label{eq:r2}
\end{equation}
where $\bar{y}$ is the mean of the observed values.
$R^2 = 1$ indicates a perfect fit; $R^2 = 0$ means the model performs
no better than predicting the mean.

%----------------------------------------------------------
\subsection{Experimental Results}
%----------------------------------------------------------

\subsubsection{Individual Model Performance}

Table~\ref{tab:model_results} reports the held-out test RMSE and $R^2$
for all six models under the best-per-model historical run.
Three models achieved their best result under the 50-feature fixed-component
PCA (run \texttt{20260325}), while the remaining three improved after
switching to the 92-feature dynamic-threshold PCA (run \texttt{20260403}).

\begin{table}[htbp]
  \centering
  \caption{Individual model performance (80/20 temporal split, best historical run per model).}
  \label{tab:model_results}
  \begin{tabular}{lccc}
    \toprule
    Model         & Test RMSE & Test $R^2$ & \#\ Features \\
    \midrule
    LightGBM      & 1.2094  & 0.9638 & 50 \\
    XGBoost       & 1.3058  & 0.9578 & 92 \\
    CatBoost      & 1.3111  & 0.9575 & 50 \\
    Elastic Net   & 1.5153  & 0.9432 & 92 \\
    Lasso         & 1.6224  & 0.9348 & 92 \\
    Random Forest & 1.7702  & 0.9224 & 50 \\
    \bottomrule
  \end{tabular}
\end{table}

The three gradient boosting methods (LightGBM, CatBoost, XGBoost) consistently
outperform the linear and Random Forest baselines on the test set.
LightGBM achieves the lowest test RMSE of \textbf{1.2094} ($R^2=0.9638$).
The dynamic-threshold PCA brought notable improvements to the linear models:
Elastic Net dropped from 1.7230 to \textbf{1.5153} and Lasso from 1.7284
to \textbf{1.6224}, as the richer 92-feature representation captures
more variance that linear methods can exploit directly.
XGBoost also improved from 1.3767 to \textbf{1.3058}.

\subsubsection{Stacking Ensemble Performance}

\begin{table}[htbp]
  \centering
  \caption{Stacking ensemble results (Top-3 base models: LightGBM, CatBoost, XGBoost).}
  \label{tab:ensemble_results}
  \begin{tabular}{lccc}
    \toprule
    Method              & Test RMSE & Test $R^2$ & Weights \\
    \midrule
    LightGBM (best single)        & 1.2094 & 0.9638 & -- \\
    CatBoost                      & 1.3111 & 0.9575 & -- \\
    XGBoost                       & 1.3767 & 0.9531 & -- \\
    \midrule
    Stacking (Convex SLSQP)       & \textbf{1.1993} & \textbf{0.9644} & 0.51 / 0.49 / 0.01 \\
    \bottomrule
  \end{tabular}
\end{table}

The stacking ensemble uses five candidate meta-fusion strategies---constrained
convex combination (SLSQP), inverse-OOF-RMSE$^2$ weighting, RidgeCV,
non-negative Lasso, and OLS---and automatically selects the one with the
lowest out-of-fold RMSE.
The convex combination was chosen, assigning nearly all weight to
LightGBM (0.51) and CatBoost (0.49) while effectively discarding XGBoost.
The resulting test RMSE of \textbf{1.1993} ($R^2=0.9644$) improves upon the
best single model by 0.8\%.